Can Auxiliary Indicators Improve COVID-19 Forecasting and Hotspot Prediction?

### 2021-06-17 forecast/paper/paper.tex
drafted udf and lead/lag sections of the manuscript

\subsection{Signals and Locations}

@@

Recall $Y_{\ell,t}$ denotes the 7-day trailing average of COVID-19 case
incidence rates in location $\ell$ and at time $t$.  Separately for each
$a=7,\ldots,21$, to predict $Y_{\ell,t+a}$ for ahead value $a$, we consider a
simple probabilistic forecasting model of the form:     
\begin{equation}
\label{eq:forecast_ar}
\Quantile_\tau(Y_{\ell,t+a} \,|\, Y_{\ell,s}, s \leq t)  
= \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j}.  
\end{equation}
This model uses current case rates, and the case rates 7 and 14 days ago, in
order to predict (the quantiles of) case rates in the future.  We consider a
total of 7 quantile levels (chosen in accordance with the county-level quantile
levels suggested by the COVID-19 Forecast Hub),      
\begin{equation}
\label{eq:quantile_levels}
\tau \in \{0.025, 0.1, 0.25, 0.5, 0.75, 0.9, 0.975 \}.
\end{equation}
We fit \eqref{eq:forecast_ar} using \textit{quantile regression}
\cite{Koenker:1978, Koenker:2005, Koenker:2006} separately for each $\tau$, 
using data from all 306 HRRs, and within each HRR, using the most recent 21 days
of training data.  This gives us 6,426 training samples for each quantile 
regression problem.

@@

We treat hotspot prediction as a binary classification problem and use a setup 
altogether quite similar to the forecasting setup described previously.
Separately for each $a=7,\ldots,21$, to predict $Z_{\ell,t+a}$, we consider a
simple logistic model:
\begin{equation}
\label{eq:hotspot_ar}
\logit \big( \P(Z_{\ell,t+a} = 1 \,|\, Y_{\ell,s}, s \leq t) \big) 
= \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y^\Delta_{\ell,t-7j},
\end{equation}
where $\logit(p) = \log(p/(1-p))$, the log-odds of $p$.

@@

\subsection{Performance in Up, Down, Flat Periods}

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{fig/upswing-histogram-1.pdf}
  \caption{Histogram of the difference in WIS for the AR model relative to each
    forecaster. Larger values are better for the indicator assisted model.}
  \label{fig:up-down-flat}
\end{figure}

@@

@@

Can auxiliary indicators improve COVID-19 forecasting and hotspot prediction
models?  Our answer, based on analyzing five auxiliary indicators from the  
COVIDcast API (defined using from medical insurance claims, internet-based
surveys, and Google search trends) is undoubtedly ``yes''. However, there are 
levels of nuance to such an answer that must be explained.  None of the
indicators that we have investigated appear to be the ``silver bullet'' that 
one might have hoped for, revolutionizing the tractability of the prediction 
problem, rendering it easy when it was once hard (in the absence of auxiliary 
information).  Rather, the gains in accuracy from the indicator models (over an
autoregressive model based only on past case rates) appear to be nontrivial, and
consistent across modes of analysis, but modest.  In forecasting, the indicator 
models appear to be most useful in periods in which case rates are flat or
trending down, rather than periods in which case rates are trending up (as one 
might hope to see is the benefit provided by a hypothetical ``leading
indicator'').

As mentioned previously, the indicator models could likely be improved by using 
location-specific effects, as well as using nowcasting techniques to estimate
finalized indicator values before using them as features (in order to account
for backfill in the claims-based signals in particular).  Beyond this, it is
certainly possible that more sophisticated models for forecasting or hotspot
prediction would lead to different results, and even different
insights. Natural directions to explore include using interactions, using
multiple indicators in a single model, and leveraging HRR demographics or
mobility patterns.  That said, we are doubtful that more sophisticated modeling
techniques would change the ``topline'' conclusion---that auxiliary indicators 
can provide nontrivial, consistent, but modest gains in forecasting and
hotspot prediction. Whether a more sophisticated model would be able to leverage
the indicators in such a way as to change some of the finer conclusions---e.g.,
by offering clear improvements in periods in which cases are trending up---is 
less clear to us.

>>>>

\subsection{Signals and Locations}
\label{sec:signals-locations}

@@

Recall $Y_{\ell,t}$ denotes the 7-day trailing average of COVID-19 case
incidence rates in location $\ell$ and at time $t$.  Separately for each
$a=7,\ldots,28$, to predict $Y_{\ell,t+a}$ for ahead value $a$, we consider a
simple probabilistic forecasting model of the form:     
\begin{equation}
\label{eq:forecast_ar}
\Quantile_\tau(Y_{\ell,t+a} \,|\, Y_{\ell,s}, s \leq t)  
= \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j}.  
\end{equation}
This model uses current case rates, and the case rates 7 and 14 days ago, in
order to predict (the quantiles of) case rates in the future.  We consider a
total of 7 quantile levels (chosen in accordance with the county-level quantile
levels suggested by the COVID-19 Forecast Hub),      
\begin{equation}
\label{eq:quantile_levels}
\tau \in \{0.025, 0.1, 0.25, 0.5, 0.75, 0.9, 0.975 \}.
\end{equation}
We fit \eqref{eq:forecast_ar} using \textit{quantile regression}
\cite{Koenker:1978, Koenker:2005, Koenker:2006} separately for each $\tau$, 
using data from all 306 HRRs, and within each HRR, using the most recent 21 days
of training data.  This gives us 6,426 training samples for each quantile 
regression problem.

@@

We treat hotspot prediction as a binary classification problem and use a setup 
altogether quite similar to the forecasting setup described previously.
Separately for each $a=7,\ldots,28$, to predict $Z_{\ell,t+a}$, we consider a
simple logistic model:
\begin{equation}
\label{eq:hotspot_ar}
\logit \big( \P(Z_{\ell,t+a} = 1 \,|\, Y_{\ell,s}, s \leq t) \big) 
= \alpha^{a,\tau} + \sum_{j=0}^2 \^{a,\tau}_j Y^\Delta_{\ell,t-7j},
\end{equation}
where $\logit(p) = \log(p/(1-p))$, the log-odds of $p$.

@@

\subsection{Performance in Up, Down, and Flat Periods}

The course of the pandemic has played out very differently across space and
time. Aggregating case rates nationally shows three pronounced waves, but the
behavior is more nuanced at the HRR level. 
Figure \ref{fig:trajectory} is a single example of what a forecast looks like
during a period of relatively flat cases as New York City begins what will
become its second wave. The AR forecast's 50\% confidence interval contains this
upswing, but the median forecast is clearly below the finalized data.
Unfortunately, this behavior is typical of all forecasters: during upswings, the
median forecast tends to fall below the target while the reverse is true during
downswings.

Figure \ref{fig:up-down-flat} shows histograms of the difference between the WIS for the
forecasters with the benefit of the indicators and the WIS of the AR model, but
it stratifies these differences by
whether the target occurs during a period of increasing cases rates (up),
decreasing case rates (down), or flat. To define the increasing period, we use
the same definition we used for the hotspot task in Table
\ref{tab:analysis_tasks}. Thus all hotspots are ``up'' while all non-hotspots
are either flat or decreasing. For the ``down'' scenario, we simply use the opposite of
the hotspot definition: $Y_{\ell,t}$ decreases by more than 25\% relative to the
preceding week.

While the performance of all forecast degrades during up periods (WIS is on the
scale of cases, so higher case counts means higher WIS) there are pronounced
patterns for the different models. 
Google-AA is the only model that
significantly outperforms the AR model during upswings. Both Google-AA and
CTIS-CLI-in-community have large right tails during the flat periods while the
claims signals have large left tails (indicating some large, relatively poor
forecasts) during flat and up periods. CHNG-CLI, CHNG-COVID, Google-AA and
especially CTIS-CLI-in-community have large positive outliers during the down
period. Overall, the indicators seem to help more during periods of flat or down
cases than during up periods, with the exception of Google-AA. In the
supplement, we examine classification accuracy and log-likelihood for the
hotspot task. We see a similar phenomenon: the indicators dramatically improve
hotspot prediction accuracy during flat and down periods with more mixed
behavior during periods of increasing cases, with CHNG-CLI, CHNG-COVID, and DV-CLI
leading to decreased performance.

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{fig/upswing-histogram-1.pdf}
  \caption{Histogram of the difference in WIS for the AR model relative to each
    forecaster. Larger values are better for the indicator assisted model. The
    $y$-axis is on the logarithmic scale to 
    emphasize outliers.}
  \label{fig:up-down-flat}
\end{figure}

@@

As described in Section \ref{sec:signals-locations}, each of the indicators we
examine could be said to measure aspects on disease progression that would
precede a positive test. That is, we imagine that these signals should ``lead''
cases. It is entirely reasonable to imagine that, prior to an increase of
confirmed COVID-19 tests reported by public health in a particular location, we
would see an increase in insurance claims for COVID-19-related medical visits.
However, it may well be the case that this leading behavior is different during
different periods. In fact, we find that the ``leadingness'' of the indicators
is more pronounced during downswings and flat periods than during upswings, a
plausible explanation for the decreased performance noted above.

In order to measure the leadingness on an indicator, we use the cross
correlation function. Essentially this examines the Pearson correlation between a
shifted version of the indicator and the case rate. So if $X_{\ell,t}$ is
shifted forward by 7 days, and the magnitude of the correlation of that shifted series with
$Y_{\ell,t}$ (unshifted) is high, we would say that $X_{\ell}$ leads $Y_{\ell}$
by 7 days. This process is repeated for a range of shifts, forward and backward,
resulting in leadingness and laggingness scores. Further details are given in
the supplement.

Figure \ref{fig:leading-lagging} shows the correlations between the difference
in the WIS of the AR model relative to $F$ with the leading and laggingness
scores stratified by whether the target is classified as up, down, or flat. One
would expect that increases in $\WIS(AR) - \WIS(AR)$ would be positively
correlated with leadingness. This relationship is strongest during down periods
and weakest during up periods with flat in the middle. In fact, for each
indicator, the strength of correlations between improved performance and
leadingness can be ordered down $>$ flat $>$ up. The relationship with
laggingness is the opposite, though less pronounced. CHNG-CLI's forecast
performance doesn't correlate very strongly with laggingness while any
laggingness in Google-AA during down and flat periods results in realtively
worse forecasting performance.

Because a time series can, to a degree, both simultaneously leading and lagging,
we also examine the correlation between performance and the difference between
leadingness and laggingness. Figure \ref{fig:diff-btw-leading-lagging} shows
this behavior. Again, indicators that are more leading then lagging tend to
are positively correlated with improved relative performance. This relationship
is most pronounced during periods of decreasing case rates and attenuates in
other scenarios.

@@

Can auxiliary indicators improve COVID-19 forecasting and hotspot prediction
models?  Our answer, based on analyzing five auxiliary indicators from the  
COVIDcast API (defined using from medical insurance claims, internet-based
surveys, and Google search trends) is undoubtedly ``yes''. However, there are 
levels of nuance to such an answer that must be explained.  None of the
indicators that we have investigated appear to be the ``silver bullet'' that 
one  might have hoped for, revolutionizing the tractability of the prediction 
problem, rendering it easy when it was once hard (in the absence of auxiliary 
information).  Rather, the gains in accuracy offered by the indicator models 
(over an autoregressive model based on past case rates along) appear to be
nontrivial, and consistent across modes of analysis, but modest.  In
forecasting, the indicator models appear to be most useful in periods in which 
case rates are flat or trending down, rather than periods in which case rates
are trending up (as one might hope to see, as the hypothetical benefit provided
by an idealized ``leading indicator'').

As mentioned previously, the indicator models could certainly benefit from a 
careful handling of location-specific effects (to account for spatial
heterogeneity in the search trends and claims-based indicators in particular),
as well as nowcasting techniques to estimate finalized indicator values before
using them as features (to account for backfill in the claims-based indicators
in particular).  Beyond this, it is certainly possible that more sophisticated 
models for forecasting or hotspot prediction would lead to different results,
and potentially even different insights.  Natural directions to explore include  
using interactions, using multiple indicators in tandem in a single model, and 
leveraging HRR demographics or mobility patterns.  That said, we are doubtful
that the ``topline'' conclusion would change--that auxiliary indicators can
provide a nontrivial, consistent, but modest improvement in forecasting and
hotspot prediction.

Furthermore, from our experience working on forecasting  TODO
